### 2022-05-13 inst/manuscript.Rmd
packages used
fixes from the proof reading

**KEYWORDS**: Eutrophication, Pinheiros river, Urban Reservoir.

@@

Eutrophication is enrichment by nutrients (mainly nitrogen and phosphorus) in aquatic ecosystems and constitutes a growing global problem [@sinhaEutrophicationWillIncrease2017].  Eutrophication causes a decrease in water quality and prevent its use for human consumption in many cases [@nogueiraEutrofizacaoNoReservatorio2015; @andradeBalancoNitrogenioFosforo2020; @nobrePrecipitationLandscapeProperties2020]. For example, eutrophication increases turbidity and algae biomass, particularly toxin-producing groups (e.g., *Microcystis aeruginosa*).

@@

The analysis and visualization of data and statistical analysis were performed using the R software (*https://www.r-project.org/*). R Packages used in this research are available on the Comprehensive R Archive Network (CRAN).

@@

Water temperature (WT), dissolved oxygen (DO), pH and electrical conductivity (EC) of water were measured *in situ* using a HACH sensor (model HQ40D).
Water samples were collected and analyzed for concentration of parameters: Total nitrogen (TN) [@valderramaSimultaneousAnalysisTotal1981], Nitrate (NO3) [@mackereth1978water], Nitrite (NO2) [@mackereth1978water], Ammonium (NH4) [@koroleffDeterminationNutrients1976], total phosphorus (TP) [@valderramaSimultaneousAnalysisTotal1981], total dissolved phosphorus (TDP) [@stricklandManualSeawaterAnalysis1960], Soluble reactive phosphorus (SRP) [@stricklandManualSeawaterAnalysis1960], and Chlorophyll-a (CHLA) [@cetesbNormaTecnicaL52014].
The Trophic State Index for tropical/subtropical reservoirs (TSI-tsr) was calculated according to @CUNHA2013126. The analysis and visualization of data was performed using the packages adespatial, SciViews, and tidyverse [@tidyverse2019].

@@

A principal component analysis (PCA) was performed to summarize the variation of  nutrients (TN, NO2, NO3, NH4, TP, TDP, SRP), chlorophyll-a, and the following measures from water surface: water temperature, pH, dissolved oxygen and electrical conductivity. PCA was conducted with normalized variables and with a correlation matrix [@legendrelegendreNumericalEcology2012]. Axes were retained for interpretation following the Broken-Stick criterion [@jacksonStoppingRulesPrincipal1993]. Variables with loadings higher than |0.6| were considered as relevant for PCA axes ordination. PCA was performed using R software with stats and vegan packages, and visualization of PCA ordinations were made with ggplot2 [@tidyverse2019] and patchwork packages.

@@

To explore which parameters varied according to sampling points and sampling periods, univariate STIs were performed with each physicochemical variable transformed by log10 (x + 1) (except pH) as response variable. Significance in STI was assessed using 999 permutations. A significance level of 5% was adopted and all analyses were performed in the R software with adespatial package.

@@

The Billings reservoir was classified by the State Government of São Paulo as a Class II water body',  destaque_rev2)` [@saopaulostategovernmentDecretoNo101977]. According to the aforementioned resolution, Class II waters can be used for human consumption (after conventional treatment), for protection of aquatic communities, for primary contact recreation, for irrigation of vegetables, fruit plants and parks, gardens, sports fields and leisure, with which the public may have direct contact, aquaculture, and fishing activities',  destaque_rev2)` [@nationalenvironmentcouncilRESOLUCAOCONAMANo2005].

@@

```{r table-pca, tab.id = "table-pca"}
resultados_pca %>%
  dplyr::mutate(
    Variables = dplyr::case_when(
      Variables == "water_temperature" ~ "WT",
      Variables == "dissolved_oxygen" ~ "DO",
      Variables == "electrical_conductivity" ~ "EC",
      Variables == "ph" ~ "pH",
      is.na(Variables) ~ " ",
      TRUE ~ Variables
    )
  ) %>%
  dplyr::left_join(variable_names, by = c("Variables" = "variable")) %>%
  dplyr::mutate(Variables = dplyr::if_else(
    !is.na(variable_name),
    glue::glue("{variable_name} ({Variables})"),
    ""
  )) %>% 
  dplyr::rename("Group" = "group") %>% 
  dplyr::select(-variable_name) %>% 
  flextable::flextable() %>%
  flextable::merge_v(j = c("Group")) %>%
  flextable::autofit() %>% 
  flextable::fontsize(size = 9) %>%
  ftExtra::colformat_md() %>% 
  flextable::theme_zebra() %>% 
   ftExtra::span_header() %>% 
  flextable:: align(align = "center", part = "all") %>% 
  flextable::width(j = "Group", width = 1) %>% 
  flextable::merge_at(i = 1:1, j = 1:2 ) %>% 
  flextable::merge_at(i = 2:2, j = 1:2 ) %>% 
  flextable::set_caption(
    "Loadings, Eigenvalues and Proportion of explained variance of a Principal Component Analysis (PCA) ordination summarizing limnological variables sampled in Billings reservoir from August, 2016 to February, 2019. Only the axes with eigenvalues higher than those expected by a Broken-Stick distribution are presented. Variables with loadings higher than |0.6| are highlighted in bold."
  )

@@

Results for Total dissolved phosphorus (TDP) ranged from `r dplyr::pull(stats_tdp_dry, min)` to `r dplyr::pull(stats_tdp_dry, max)` µg/L during the dry season and from  `r dplyr::pull(stats_tdp_rainy, min)` to `r dplyr::pull(stats_tdp_rainy, max)` µg/L during the rainy season. The soluble reactive phosphorus (SRP) fractionor Orthophosphate comprises all types of inorganic phosphorus in solution [@markadTrophicStateModeling2019]. SRP varied seasonally, being higher and more variable in the dry season than the rainy season for all sampling points. SRP results in the dry season ranged from `r dplyr::pull(stats_srp_dry, min)` to `r dplyr::pull(stats_srp_dry, max)` µg/L, and from `r dplyr::pull(stats_srp_rainy, min)` to `r dplyr::pull(stats_srp_rainy, max)` µg/L during the rainy season. As SRP is the main form of phosphorus assimilated by phytoplankton, the lower values obtained in the rainy season can be associated with greater phytoplankton biomass (Chlorophyll-a) obtained in that season.

@@

The first author thanks the São Paulo Research Foundation (FAPESP) - Grant number XXXX/XXXXX-X. The third author thanks the Coordination for the Improvement of Higher Level Personnel (CAPES) for a postdoctoral scholarship. The last author thanks the  São Paulo Research Foundation (FAPESP) and Basic Sanitation Utility Company of the State of São Paulo (SABESP) - Grant number XXXX/XXXXX-X.

>>>>

**KEYWORDS**: Eutrophication, Pinheiros river, Urban reservoir.

@@

Eutrophication is the enrichment by nutrients (mainly nitrogen and phosphorus) in aquatic ecosystems and constitutes a growing global problem [@sinhaEutrophicationWillIncrease2017].  Eutrophication causes a decrease in water quality and prevent its use for human consumption in many cases [@nogueiraEutrofizacaoNoReservatorio2015; @andradeBalancoNitrogenioFosforo2020; @nobrePrecipitationLandscapeProperties2020]. For example, eutrophication increases turbidity and algae biomass, particularly toxin-producing groups (e.g., *Microcystis aeruginosa*).

@@

The analysis and visualization of data and statistical analysis were performed using the R software (*https://www.r-project.org/*) [@R-software]. R Packages used in this research are available on the Comprehensive R Archive Network (CRAN).

@@

Water temperature (WT), dissolved oxygen (DO), pH and electrical conductivity (EC) of water were measured *in situ* using a HACH sensor (model HQ40D).
Water samples were collected and analyzed for concentration of parameters: Total nitrogen (TN) [@valderramaSimultaneousAnalysisTotal1981], Nitrate (NO3) [@mackereth1978water], Nitrite (NO2) [@mackereth1978water], Ammonium (NH4) [@koroleffDeterminationNutrients1976], total phosphorus (TP) [@valderramaSimultaneousAnalysisTotal1981], total dissolved phosphorus (TDP) [@stricklandManualSeawaterAnalysis1960], Soluble reactive phosphorus (SRP) [@stricklandManualSeawaterAnalysis1960], and Chlorophyll-a (CHLA) [@cetesbNormaTecnicaL52014].
The Trophic State Index for tropical/subtropical reservoirs (TSI-tsr) was calculated according to @CUNHA2013126. The analysis and visualization of data was performed using the packages adespatial [@R-adespatial], SciViews [@R-SciViews], and tidyverse [@tidyverse2019].

@@

A principal component analysis (PCA) was performed to summarize the variation of  nutrients (TN, NO2, NO3, NH4, TP, TDP, SRP), chlorophyll-a, and the following measures from water surface: water temperature, pH, dissolved oxygen and electrical conductivity. PCA was conducted with normalized variables and with a correlation matrix [@legendrelegendreNumericalEcology2012]. Axes were retained for interpretation following the Broken-Stick criterion [@jacksonStoppingRulesPrincipal1993]. Variables with loadings higher than |0.6| were considered as relevant for PCA axes ordination. PCA was performed using R software [@R-software] with stats and vegan [@R-vegan] packages, and visualization of PCA ordinations were made with ggplot2 [@tidyverse2019] and patchwork packages.

@@

To explore which parameters varied according to sampling points and sampling periods, univariate STIs were performed with each physicochemical variable transformed by log10 (x + 1) (except pH) as response variable. Significance in STI was assessed using 999 permutations. A significance level of 5% was adopted and all analyses were performed in the R software [@R-software] with adespatial [@R-adespatial] package.

@@

The Billings reservoir was classified by the State Government of São Paulo as a Class II water body', [@saopaulostategovernmentDecretoNo101977]. According to the aforementioned resolution, Class II waters can be used for human consumption (after conventional treatment), for protection of aquatic communities, for primary contact recreation, for irrigation of vegetables, fruit plants and parks, gardens, sports fields and leisure, with which the public may have direct contact, aquaculture, and fishing activities',  destaque_rev2)` [@nationalenvironmentcouncilRESOLUCAOCONAMANo2005].

@@

```{r table-pca, tab.id = "table-pca"}
resultados_pca %>%
  dplyr::mutate(
    Variables = dplyr::case_when(
      Variables == "water_temperature" ~ "WT",
      Variables == "dissolved_oxygen" ~ "DO",
      Variables == "electrical_conductivity" ~ "EC",
      Variables == "ph" ~ "pH",
      is.na(Variables) ~ " ",
      TRUE ~ Variables
    )
  ) %>%
  dplyr::left_join(variable_names, by = c("Variables" = "variable")) %>%
  dplyr::mutate(Variables = dplyr::if_else(
    !is.na(variable_name),
    glue::glue("{variable_name} ({Variables})"),
    ""
  )) %>% 
  dplyr::rename("Group" = "group") %>% 
  dplyr::select(-variable_name) %>% 
  flextable::flextable() %>%
  flextable::merge_v(j = c("Group")) %>%
  flextable::autofit() %>% 
  flextable::fontsize(size = 9) %>%
  ftExtra::colformat_md() %>% 
  flextable::theme_zebra() %>% 
   ftExtra::span_header() %>% 
  flextable:: align(align = "center", part = "all") %>% 
  flextable::width(j = "Group", width = 1) %>% 
  flextable::merge_at(i = 1:1, j = 1:2 ) %>% 
  flextable::merge_at(i = 2:2, j = 1:2 ) %>% 
  flextable::set_caption(
    "Loadings, Eigenvalues and Proportion of explained variance of a Principal Component Analysis (PCA) ordination summarizing limnological variables sampled in Billings reservoir from August, 2016 to February, 2019. Only the axes with eigenvalues higher than those expected by a Broken-Stick distribution are presented."
  )

@@

Results for Total dissolved phosphorus (TDP) ranged from `r dplyr::pull(stats_tdp_dry, min)` to `r dplyr::pull(stats_tdp_dry, max)` µg/L during the dry season and from  `r dplyr::pull(stats_tdp_rainy, min)` to `r dplyr::pull(stats_tdp_rainy, max)` µg/L during the rainy season. The soluble reactive phosphorus (SRP) fraction or Orthophosphate comprises all types of inorganic phosphorus in solution [@markadTrophicStateModeling2019]. SRP varied seasonally, being higher and more variable in the dry season than the rainy season for all sampling points. SRP results in the dry season ranged from `r dplyr::pull(stats_srp_dry, min)` to `r dplyr::pull(stats_srp_dry, max)` µg/L, and from `r dplyr::pull(stats_srp_rainy, min)` to `r dplyr::pull(stats_srp_rainy, max)` µg/L during the rainy season. As SRP is the main form of phosphorus assimilated by phytoplankton, the lower values obtained in the rainy season can be associated with greater phytoplankton biomass (Chlorophyll-a) obtained in that season.

@@

The first author thanks the São Paulo Research Foundation (FAPESP) - Grant number 	18/23771-6. The third author thanks the Coordination for the Improvement of Higher Level Personnel (CAPES) for a postdoctoral scholarship - Grant Number 88887.374969/2019-00. The last author thanks the  São Paulo Research Foundation (FAPESP) and Basic Sanitation Utility Company of the State of São Paulo (SABESP) - Grant number PITE-SABESP 2010/50738-8.
